Interisland movements of woodland caribou in Newfoundland

### 2020-02-19 paper/manuscript.Rmd
calving swim-bo-baggins

Predator avoidance could contribute to swimming behaviour of caribou (Miller
2002). Although caribou co-evolved with wolves (*Canis lupus*) throughout their range, wolves are extirpated from Newfoundland, including Fogo Island. Coyotes (*Canis latrans*)
and black bears are the primary predators of
caribou (Bastille-Rousseau et al. 2016). While predation by coyotes on adult
female caribou is possible, it is relatively rare on Fogo Island (Webber &
Vander Wal unpublished data). Therefore, any effect of predation on swimming
behaviour is likely related to fine-scale encounters between coyotes and caribou
which result in caribou swimming as a flight response. Alternatively, during
calving, females may swim to small islands to give birth (Bergerud & Page 1987)
or swim to islands with a calf at heel to avoid predation. Swimming with a calf
at heel has been observed for caribou in freshwater lakes on Fogo Island (Webber
& Vander Wal unpublished data), so we anticipate it may also be possible in the
ocean. We suggest future studies assess encounter rates among caribou and
coyotes to determine what proportion of encounters result in swimming
behaviour.Woodland caribou are well known to disperse to remote locations,
including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987).
Moving to remote locations is a viable anti-predator strategy for caribou
because it reduces detection and encounter rates by predators. For instance, the
use of shoreline habitat in summer was deemed an effective anti-predator calving
strategy because caribou avoided predators by using islands and peninsulas or
swimming as a flight response (Bergerud et al. 1990). The use of water, and by
association occasionally swimming, is therefore part of the fine-scale
interactions between caribou and their predators and the use of islands appears
to be an effective anti-predator strategy. Coyotes and black bears are
responsible for the majority of mortalities for neonate caribou calves
(Bastille-Rousseau et al. 2016), although predation can still occur after the
calving period (Lewis & Mahoney 2014). However, on Fogo Island, coyotes are the
only predator of caribou.

>>>>

Predator avoidance could contribute to swimming behaviour of caribou (Miller
2002). Although caribou co-evolved with wolves (*Canis lupus*) throughout their
range, wolves are extirpated from Newfoundland, including Fogo Island. Coyotes
(*Canis latrans*) and black bears (*Ursus americanus*) are the primary predators
of caribou in Newfoundland (Bastille-Rousseau et al. 2016). While predation by
coyotes on adult female caribou does occur (ref?), it is relatively rare and the
effect of predation on swimming behaviour is more likely the result of
encounters which result in a flight response. For example, during calving,
females may swim to small islands to give birth (Bergerud & Page 1987) or swim
to islands in freshwater lakes with their calves to avoid predation.

Woodland caribou are also well known to disperse to remote locations,
including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987).
Moving to remote locations is a viable anti-predator strategy for caribou
because it reduces detection and encounter rates by predators. For instance, the
use of shoreline habitat in summer was deemed an effective anti-predator calving
strategy because caribou avoided predators by using islands and peninsulas or
swimming as a flight response (Bergerud et al. 1990). The use of water, and by
association occasionally swimming, is therefore part of the fine-scale
interactions between caribou and their predators and the use of islands appears
to be an effective anti-predator strategy. Coyotes and black bears are
responsible for the majority of mortalities for neonate caribou calves
(Bastille-Rousseau et al. 2016), although predation can still occur after the
calving period (Lewis & Mahoney 2014). However, on Fogo Island, coyotes are the
only predator of caribou.
